High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-01-28 PNAS_manuscript/PNAS_manuscript.tex

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

@@

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Yiming Zhang}
\author[a]{Nicholas L. Swanson-Hysell} 
\author[a,b]{Margaret S. Avery}
\author[c]{Roger R. Fu}

@@

% Please give the surname of the lead author for the running footer
\leadauthor{Zhang}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from the remanent magnetization of rocks is crucial for constraining the long-term evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal rock magnetic behavior. In this study, we use plagioclase cumulates (anorthosite) that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These anorthosite xenoliths have experienced minimal alteration since their emplacement and yield high quality paleointensity estimates. In contrast to predictions of a progressively decaying field leading up to late nucleation of Earth's inner core, these data record a strong virtual dipole moment 1.1 billion years ago. A strong field persisted over at least a 20-million-year interval indicating the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic.}

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$140 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths requires the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 20 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon and could challenge the hypothesis of a young inner core.
\end{abstract}

@@

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref.  \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

@@

Following IZZI-modified paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase site-level results pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The paleointensity quality index (Q$_{PI}$;  \citealp{Biggin2014a}) for the anorthosite xenoliths are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a mean of 38.28 $\pm$ 11.92 $\mu$T. This paleointensity value corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthorg/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

@@

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy (Fig. \ref{fig:coercivity}). A compilation of median destructive field (MDF) values in Fig. \ref{fig:coercivity} shows that specimens from anorthosites that pass paleointensity selection criteria have distinctly higher average MDF values than other anorthosite and diabase specimens. Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that a population of magnetic grains with more multidomain-like behavior is responsible for the non-ideal paleointensity behaviors during experiments on such specimens.

@@

We applied full TRMs to a set of anorthosite specimens with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that this set of anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions ( \citealp{Selkin2007a}; Fig. S4).

@@

The high paleointensity estimates from the 1.1 billion-year-old Midcontinent Rift rocks challenge the hypothesized monotonic decay of the strength of the geomagnetic field throughout the Proterozoic Era. Instead, the well-preserved ca. 1092 Ma anorthosite xenoliths of the Beaver River diabase record a strong geomagnetic field in the late Mesoproterozoic that exceeds the strength of the modern-day field for which crystallization of the inner core is a power source (Fig. \ref{fig:PINT_compilation}). Together with previous records obtained from the ca. 1106 Ma Osler Volcanics of the Midcontinent Rift \cite{Sprain2018a}, these data indicate that appreciable power to Earth's dynamo persisted through at least 14 Myr during the late Mesoproterozoic to maintain a strong surface field (Fig. \ref{fig:PINT_compilation}). In addition to these high geomagnetic fields recorded by Midcontinent Rift rocks, the ca. 755 Ma Mundine Well dikes \cite{Lloyd2021b} also require a stronger geomagnetic field in the Neoproterozoic than would be predict by a progressive Proterozoic decline (Fig. \ref{fig:PINT_compilation}).

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the Abitibi dikes paleointensity estimates do not go to values as low as the Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This timing would be broadly consistent with that proposed by \citealp{Biggin2015a}. However, a model prediction could be of sustained strong field values following inner core nucleation which is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

Given that the multiple records of a weak field can not all be a minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of such elevated core-mantle boundary heat flux conditions at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

@@

\acknow{Project research was supported by NSF CAREER grant EAR-1847277 to N.L.S.-H., an Institute on Lake Superior Geology Student Research Fund grant to Y.Z., and an Institute for Rock Magnetism Visiting Student Fellowship to Y.Z. The Institute for Rock Magnetism is supported by the Instrumentation and Facilities program of the NSF Earth Science Division. Permits for fieldwork and sampling from the Minnesota Department of Natural Resources are gratefully acknowledged. We thank James Pierce and Blake Hodgin for their assistance in the field. We thank Jim Miller for providing us with the Duluth Complex anorthosite thin section. We thank Dario Bilardello, Peat Solheid, Mike Jackson, Josh Feinberg, and Bruce Moskowitz for their tremendous help associated with instrument operation, data interpretations, and research guidance at the IRM. Conversations with Bruce Buffett and Zachary Geballe informed our perspectives on the geodynamic and mineral physics literature. Paleomagnetic data associated with this study are available within the MagIC database (\url{https://earthorg/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{provide link when doi is generated at time of proofs}) and all data are within a Github repository associated with this work along with Python code related to calculations and visualizations associated with the study that is also archived on Zenodo (\textit{provide link when doi is generated at time of proofs}).}

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}

\templatetype{pnasresearcharticle}

@@

\author[a,1]{Yiming Zhang}
\author[a]{Nicholas L. Swanson-Hysell} 
\author[a,b]{Margaret S. Avery}
\author[c]{Roger R. Fu}

@@

\leadauthor{Zhang}

\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from the remanent magnetization of rocks is crucial for constraining the long-term evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal rock magnetic behavior. In this study, we use plagioclase cumulates (anorthosite) that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These anorthosite xenoliths have experienced minimal alteration since their emplacement and yield high quality paleointensity estimates. In contrast to predictions of a progressively decaying field leading up to late nucleation of Earth's inner core, these data record a strong virtual dipole moment 1.1 billion years ago. A strong field persisted over at least a 20-million-year interval indicating the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic.}

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$140 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths require the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 14 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon and could challenge the hypothesis of a young inner core.
\end{abstract}

@@

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref.  \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase site-level results pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The paleointensity quality index (Q$_{PI}$;  \citealp{Biggin2014a}) for the anorthosite xenoliths are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a mean of 38.28 $\pm$ 11.92 $\mu$T. This paleointensity value corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthorg/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

@@

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy (Fig. \ref{fig:coercivity}). A compilation of median destructive field (MDF) values in Figure. \ref{fig:coercivity} shows that specimens from anorthosites that pass paleointensity selection criteria have distinctly higher average MDF values than other anorthosite and diabase specimens. Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that a population of magnetic grains with more multidomain-like behavior is responsible for the non-ideal paleointensity behaviors during experiments on such specimens.

@@

We applied full TRMs to a set of anorthosite specimens with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that this set of anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (\citealp{Selkin2007a}; Fig. S4).

@@

The high paleointensity estimates from the 1.1 billion-year-old Midcontinent Rift rocks challenge the hypothesized monotonic decay of the strength of the geomagnetic field throughout the Proterozoic Era. Instead, the well-preserved ca. 1092 Ma anorthosite xenoliths of the Beaver River diabase record a strong geomagnetic field in the late Mesoproterozoic that exceeds the strength of the modern-day field for which crystallization of the inner core is a power source (Fig. \ref{fig:PINT_compilation}). Together with previous records obtained from the ca. 1106 Ma Osler Volcanics of the Midcontinent Rift \cite{Sprain2018a}, these data indicate that appreciable power to Earth's dynamo persisted through at least 14 Myr during the late Mesoproterozoic to maintain a strong surface field (Fig. \ref{fig:PINT_compilation}). In addition to these high geomagnetic fields recorded by Midcontinent Rift rocks, the ca. 755 Ma Mundine Well dikes \cite{Lloyd2021b} also require a stronger geomagnetic field in the Neoproterozoic than would be predicted by a progressive Proterozoic decline (Fig. \ref{fig:PINT_compilation}).

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the Abitibi dikes paleointensity estimates do not go to values as low as the Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This timing would be broadly consistent with that proposed by \citealp{Biggin2015a}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

Given that the multiple records of a weak field can not all be a minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of such elevated core-mantle boundary heat flux conditions at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby also explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions (Fig. \ref{fig:PINT_compilation}; \citealp{Bono2019a}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

@@

\acknow{Project research was supported by NSF CAREER grant EAR-1847277 to N.L.S.-H., an Institute on Lake Superior Geology Student Research Fund grant to Y.Z., and an Institute for Rock Magnetism Visiting Student Fellowship to Y.Z. The Institute for Rock Magnetism is supported by the Instrumentation and Facilities program of the NSF Earth Science Division. Permits for fieldwork and sampling from the Minnesota Department of Natural Resources are gratefully acknowledged. We thank James Pierce and Blake Hodgin for their assistance in the field. We thank Jim Miller for providing us with the Duluth Complex anorthosite thin section. We thank Dario Bilardello, Peat Solheid, Mike Jackson, Josh Feinberg, and Bruce Moskowitz for their tremendous help associated with instrument operation, data interpretations, and research guidance at the IRM. Conversations with Bruce Buffett and Zachary Geballe informed our perspectives on the geodynamic and mineral physics literature. Paleomagnetic data associated with this study are available within the MagIC database (\url{https://earthorg/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{provide link when doi is generated at time of proofs}) and all data are within a Github repository along with Python code related to calculations and visualizations (\url{https://github.com/Swanson-Hysell-Group/AX_BD_PINT}) that are also archived on Zenodo (\textit{provide link when doi is generated at time of proofs}).}
